\ \\ \LARGE\bf Generating Lambda Term Individuals in Typed Genetic Programming Using Forgetful A*

### 2014-01-21 tex/cec14b/article.tex
cecek ... resubmitly-finale

~\\
Furthermore, our system utilizes generalization of the standard 
tree-swapping crossover operator. Since it is beyond the scope of this 
paper we mention it only briefly. Two main concerns with swapping typed subtrees 
are types and free variables. Well-typed offspring is obtained 
by swapping only subtrees of the same type. Only subtrees with 
corresponding counterpart in the second parent are randomly chosen from.
More interesting problem lies in free variables, which may cause trouble
if swapped somewhere where it is suddenly not bounded. In order to circumvent this
difficulty we utilize technique called \textit{abstraction elimination}\cite{jones87}
that transforms an arbitrary \lterm into \lterm that contains no lambda abstractions
and no bound variables. After the initial population is generated,  it is transformed
by abstraction elimination. Another possible transformation taking place
after initialization is $\eta$-normalization shortening 
rather long long normal form into $\beta\eta$-normal form.  
Another performance enhancing transformation is use of "applicative" tree 
representation (coming directly from inductive definition of \lterms).

@@

\subsection{Simple symbolic regression}
\textit{Simple Symbolic Regression} is a problem described
in \cite{koza92}. Objective of this problem is to 
find a function $f(x)$ that fits a sample
of twenty given points. The target function is 
function $f_{t}(x) = x^4 + x^3 + x^2 + x$.  
Desired type of generated programs $\tau$ and 
building blocks context $\Gamma$ are following.
\begin{align*}
\tau = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,
  (-)   : \Real \ar \Real \ar \Real    ,
  (*)   : \Real \ar \Real \ar \Real    ,\\
  rdiv &: \Real \ar \Real \ar \Real    ,
  sin   : \Real \ar \Real              ,
  cos   : \Real \ar \Real              ,\\
  exp  &: \Real \ar \Real              , 
  rlog  : \Real \ar \Real              \}
\end{align*}
where

@@

~\\Fitness function is computed as follows.

$$ fitness(f) =  \frac{1}{1+ \sum\limits_{i=1}^{20}{ \vert f(x_i)-y_i }\vert }  $$
where $(x_i,y_i)$ are 20 data samples from $[-1,1]$, such that $y_i = f_t(x_i)$.
An individual $f$ such that $\vert f(x_i)-y_i \vert < 0.01 $ for all data samples is 
considered as a correct individual.

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.14]{imgs/cec/SSR.PNG}
  \caption{Graphs for Simple symbolic regression.}
\end{figure}

@@

Figure~1 shows results of this experiment. 
Table~I summarizes statistical analysis
of the fitness values of the best individuals of the run.  
By conventional criteria, this difference is considered to be not statistically significant.
Standard \textit{ramped half-and-half strategy} scored 17/50 (34\%) success rate. 
Minimal $I(M,i,z)$ was in generation 23 with 192,000 individuals to be processed.
The average individual size for generation 50 was 68.8.
The experiment took 46 minutes.
Our \textit{geometric strategy} scored 21/50 (42\%) success rate. 
Minimal $I(M,i,z)$ was in generation 29 with 150,000 individuals to be processed.
The average individual size for generation 50 was 44.1.
The experiment took 26 minutes.
So for the \textit{Simple symbolic regression}
our geometric strategy is slightly more successful than 
ramped half-and-half strategy in these observed metrics.

\subsection{Artificial ant}
\textit{Artificial Ant} is another problem described
in \cite{koza92}. Objective of this problem is to 
find a control program for an artificial ant so
that it can find all food located on "Santa Fe" trail.
The Santa Fe trail lies on toroidal square grid.
The ant is in the upper left corner, facing right.
The ant is able to move forward, turn left, and sense if a food 
piece is ahead of him.

\begin{align*}
\tau = An&tAct\\
\Gamma = \{~~
  l    &: AntAct                              ,
  r    : AntAct                               ,
  m    : AntAct                               ,\\
  ifa  &: AntAct \ar AntAct \ar AntAct  ,\\
  p2   &: AntAct \ar AntAct \ar AntAct  ,\\
  p3   &: AntAct \ar AntAct \ar AntAct \ar AntAct  \}
\end{align*}

Action $l$ turns the ant left. 
Action $r$ turns the ant right.
Action $m$ moves the ant forward.
Action $ifa~x~y$ (if-food-ahead) performs action $x$ 
if a food piece is ahead of the ant,
otherwise it performs action $y$.
Action $p2~x~y$ first performs action $x$ and after it action $y$.
Action $p3~x~y~z$ first performs action $x$, 
after that action $y$ and finally $z$.
Actions $l, r$ and $m$ each take one time step to execute.
Ants action is performed over and over again until it reaches predefined
maximal number of steps.

Fitness value is equal to number of eaten food pieces.
An individual such that eats all 89 pieces of food is 
considered as a correct solution.
This limit is set to be 600 time steps\footnote{
In \cite{koza92} this limit is said to be 400 time steps.
But there is also mentioned following solution, 
which is described as correct solution:
\texttt{(ifa m (p3 l (p2 (ifa m r) (p2 r (p2 l r)))(p2 (ifa m l) m)))}.
This program needs 545
time steps; if it is given only 400 time steps, then it eats only 79 pieces
of food. Thus we use 600 time steps. 
}.

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.14]{imgs/cec/ANT.PNG}
  \caption{Graphs for Artificial ant problem.}
\end{figure}

@@

Figure~2 shows results of this experiment. 
Table~II summarizes statistical analysis
of the fitness values of the best individuals of the run.  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be extremely statistically significant.

Standard \textit{ramped half-and-half strategy} scored 7/50 (14\%) success rate. 
Minimal $I(M,i,z)$ was in generation 28 with 449,500 individuals to be processed.
The average individual size for generation 50 was 279.5.
The experiment took 265 minutes.

Our \textit{geometric strategy} scored 19/50 (38\%) success rate. 
Minimal $I(M,i,z)$ was in generation 10 with 115,500 individuals to be processed.
The average individual size for generation 50 was 94.1.
The experiment took 107 minutes.

This is a big improvement in all watched factors.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.14]{imgs/cec/EP.PNG}
  \caption{Graphs for Even parity problem.}
\end{figure}

@@

Figure~3 shows results of this experiment. 
Table~III summarizes statistical analysis
of the fitness values of the best individuals of the run.  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be 
statistically significant.

>>>>

~\\
Furthermore, our system utilizes generalization of the standard 
tree-swapping crossover operator. Since it is beyond the scope of this 
paper we mention it only briefly. Two main concerns with swapping typed 
subtrees are types and free variables. Well-typed offspring is obtained 
by swapping only subtrees of the same type. Only subtrees with 
corresponding counterpart in the second parent are randomly chosen from.
More interesting problem lies in free variables, which may cause trouble
if swapped somewhere where it is suddenly not bounded. In order to circumvent this
difficulty we utilize technique called \textit{abstraction elimination}\cite{jones87}
that transforms an arbitrary \lterm into \lterm that contains no lambda abstractions
and no bound variables. After the initial population is generated,  it is transformed
by abstraction elimination. Another possible transformation taking place
after initialization is $\eta$-normalization shortening 
rather long long normal form into shorter $\beta\eta$-normal form.  
Another performance enhancing transformation is option of using "applicative" tree representation (coming directly from inductive definition of \lterms) instead of more traditional S-expression representation.

A remarkable benefit of parameterizing the generating method 
by a search strategy taking the form of a simple filtration function
is that such a function can be expressed by functional language in a very 
economical way. Take for instance a search strategy that behaves same
as the \textit{geometric} except that when it comes to 
situation where all the elements are filtered out it
acts as the \textit{ramped half-and-half}. 
Such \textit{strategy composition}
is easily describable as a higher order function. Thus comes 
to mind the possibility of using our system to come up with new
search strategies on its own.

@@

\subsection{Simple symbolic regression}
\textit{Simple Symbolic Regression} is a problem described
in \cite{koza92}. Objective of this problem is to 
find a function $f(x)$ that fits a sample
of twenty given points. The target function is 
function $f_{t}(x) = x^4 + x^3 + x^2 + x$.  
Desired type of generated programs $\tau$ and 
building blocks context $\Gamma$ are following.
\begin{align*}
\tau = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,~ 
  (-)   : \Real \ar \Real \ar \Real    ,~\\
  (*)  &: \Real \ar \Real \ar \Real    ,~ 
  rdiv  : \Real \ar \Real \ar \Real    ,~\\
  sin  &: \Real \ar \Real              ,~ 
  cos   : \Real \ar \Real              ,~\\
  exp  &: \Real \ar \Real              ,~ 
  rlog  : \Real \ar \Real              \}
\end{align*}
where

@@

Fitness function is computed as follows.

$$ fitness(f) =  \frac{1}{1+ \sum\limits_{i=1}^{20}{ \vert f(x_i)-y_i }\vert }  $$
where $(x_i,y_i)$ are 20 data samples from $[-1,1]$, such that $y_i = f_t(x_i)$.
An individual $f$ such that $\vert f(x_i)-y_i \vert < 0.01 $ for all data samples is 
considered as a correct individual.\\

\begin{figure}[h!]
  \centering
  \caption{Graphs for Simple symbolic regression.}
  \includegraphics[scale=0.241]{imgs/cec/SSRb.PNG}
\end{figure}

@@

Figure~1 shows results of this experiment. 
Table~I summarizes statistical analysis
of the fitness values of the best individuals of the run 
(\textit{p-value} $ = 0.6038$).  
By conventional criteria, this difference is considered to be not statistically significant.
Standard \textit{ramped half-and-half strategy} scored 17/50 (34\%) success rate. 
Minimal $I(M,i,z)$ was in generation 23 with 192,000 individuals to be processed.
The average individual size for generation 50 was 68.8.
The experiment took 46 minutes.
Our \textit{geometric strategy} scored 21/50 (42\%) success rate. 
Minimal $I(M,i,z)$ was in generation 29 with 150,000 individuals to be processed.
The average individual size for generation 50 was 44.1.
The experiment took 26 minutes.
So for the \textit{Simple symbolic regression}
our geometric strategy is slightly more successful than 
ramped half-and-half strategy in these observed metrics.

\subsection{Artificial ant}
\textit{Artificial Ant} is another problem described
in \cite{koza92}. Objective of this problem is to 
find a control program for an artificial ant so
that it can find all food located on "Santa Fe" trail.
The Santa Fe trail lies on toroidal square grid.
The ant is in the upper left corner, facing right.
The ant is able to move forward, turn left, and sense if a food 
piece is ahead of him.
\begin{align*}
\tau~ = An&tAct\\
\Gamma = \{~~
  l    &: AntAct                              ,
  r    : AntAct                               ,
  m    : AntAct                               ,\\
  ifa  &: AntAct \ar AntAct \ar AntAct  ,\\
  p2   &: AntAct \ar AntAct \ar AntAct  ,\\
  p3   &: AntAct \ar AntAct \ar AntAct \ar AntAct  \}
\end{align*}

Actions $l$ and $r$ turn the ant left and right, respectively. 
Action $m$ moves the ant forward.
Action $ifa~x~y$ \mbox{(\textit{if-food-ahead})} 
performs action $x$ if a food piece is ahead of the ant,
otherwise it performs action $y$.
Actions $p2$ and $p3$ perform two and three consecutive actions, respectively.
Actions $l, r$ and $m$ each take one time step to execute.
Ants action is performed over and over again until it reaches predefined
maximal number of steps. 
Fitness value is equal to number of eaten food pieces.
An individual such that eats all 89 pieces of food is 
considered as a correct solution.
This limit is set to be 600 time steps\footnote{
In \cite{koza92} this limit is said to be 400 time steps.
But there is also mentioned following solution, 
which is described as correct solution:
\texttt{(ifa m (p3 l (p2 (ifa m r) (p2 r (p2 l r)))(p2 (ifa m l) m)))}.
This program needs 545
time steps; if it is given only 400 time steps, then it eats only 79 pieces
of food. Thus we use 600 time steps. 
}.

\begin{figure}[h!]
  \centering
  \caption{Graphs for Artificial ant problem.}
  \includegraphics[scale=0.26]{imgs/cec/ANTb.PNG}
\end{figure}

@@

Figure~2 shows results of this experiment. 
Table~II summarizes statistical analysis
of the fitness values of the best individuals of the run 
(\textit{p-value} $ < 0.0001$).  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be extremely statistically significant.
Standard \textit{ramped half-and-half strategy} scored 7/50 (14\%) success rate. 
Minimal $I(M,i,z)$ was in generation 28 with 449,500 individuals to be processed.
The average individual size for generation 50 was 279.5.
The experiment took 265 minutes.
Our \textit{geometric strategy} scored 19/50 (38\%) success rate. 
Minimal $I(M,i,z)$ was in generation 10 with 115,500 individuals to be processed.
The average individual size for generation 50 was 94.1.
The experiment took 107 minutes.
This is a big improvement in all watched factors.

@@

\begin{figure}[h!]
  \centering
  \caption{Graphs for Even parity problem.}
  \includegraphics[scale=0.26]{imgs/cec/EPb.PNG}
\end{figure}

@@

Figure~3 shows results of this experiment. 
Table~III summarizes statistical analysis
of the fitness values of the best individuals of the run 
(\textit{p-value} $ = 0.0250$).  
According to this analysis our strategy performed better;
by conventional criteria, this difference is considered to be 
statistically significant.
